Generating Random Logic Programs Using Constraint Programming

### 2020-01-02 paper/paper.tex
Counting is closer to working

\begin{itemize}
\item Heads are too restrictive: $\mathsf{P}(X, Y, X)$ should be allowed.
\item A constraint for logical equivalence. An algorithm to reduce each tree to
  some kind of normal form.
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
\item Inference options to explore. Logspace vs normal space. Symbolic vs
  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible. Restarting after each run is
  expensive. Periodic restarts could be an option.
\end{itemize}

@@

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{arguments}[i] = 0.
  \]
\end{constraint}

\begin{definition} We define an array \variable{occurrences} of $|\variables{}|$
  sets such that, for all $v \in \variables{}$, $\variable{occurrences}[v]$ is a
  subset of $\{ 0, \dots, \maxArity{} - 1\}$ that stores the indices of
  \variable{arguments} at which $v$ is located. It is defined by the
  \variable{setsIntsChanneling} constraint equivalent to the following
  condition: for $i = 0, \dots, \maxArity{} - 1$ and $v \in \variables{}$,
  \[
    \variable{arguments}[i] = v \quad \iff \quad i \in \variable{occurrences}[v].
  \]
\end{definition}

\begin{definition}
  We define a $|\variables{}| \times 2$ matrix $\mathbf{M}$ with domain $[0,
  \maxArity{}]$ such that for $v = 0, \dots, |\variables{}| - 1$,
  \begin{align*}
    \mathbf{M}_{v,0} &= |\variable{occurrences}[v]|.\\
    \mathbf{M}_{v,1} &= \begin{cases}
      \min \variable{occurrences}[v] & \text{if } \variable{occurrences}[v] \ne \emptyset\\
      0\footnotemark & \text{otherwise.}
    \end{cases}
  \end{align*}
\end{definition}
\footnotetext{This could be any fixed value in the domain.}

\begin{example}
  Let $\maxArity{} = 4$, $\mathsf{P} \in \predicates{}$, $\variables{} = \{
  X, Y, Z \}$. Then $\mathsf{P}(Z, Y, Z)$ would be represented as:
  \begin{align*}
    \variable{predicate} &= \mathsf{P},\\
    \variable{arguments} &= [Z, Y, Z, 0],\\
    \variable{arity} &= 3,\\
    \variable{occurrences} &= [\emptyset, \{ 1 \}, \{ 0, 2 \}],\\
    \mathbf{M} &= \begin{bmatrix}
      0 & 0\\
      1 & 1\\
      2 & 0
    \end{bmatrix}.
  \end{align*}
\end{example}

@@

If the clause should be disabled, $\variable{numNodes} = 1$ and
$\variable{treeValues}[0].\variable{name} = \top$.

\section{Managing Clauses}

@@

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$). If
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate},
  \]
  then $\variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i]$.
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body. % TODO: can this be turned into lexChainLessEq?

\begin{constraint}
  Adjacency matrix representation:
  \begin{align*}
    A[i][j] = 0 \iff &\nexists k: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{name} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{constraint}

@@

Example of ordered partitions:
\[
  3 = 2 + 1 = 2 + 1 = 1 + 1 + 1,
\]
so for $n=4$, the sum would have three terms.

Number of heads for a specific predicate with arity $a \in \arities{}$:
\[
  H_a = |\constants{}|^a + \sum_{v=1}^a \binom{a}{v} |\constants{}|^{a-v}
  \sum_{k=0}^{|\variables{}|-1} \binom{v-1}{k}.
\]
First, select the $v$ positions dedicated for variables. The remaining $a-v$
constants can then be filled in $|\constants{}|^{a-v}$ ways. Filling $v$
positions with $k+1$ variables in a non-decreasing manner (without skipping any
variables) can be seen as putting $k$ `bars' in the $v-1$ spaces between $v$
positions. Each bar represents switching to the next variable.

@@

>>>>

\begin{itemize}
\item Construct matrix M for variables that don't participate in the head. Have
  a variable variablesThatParticipateInTheHead that separates the set of
  variables.
\item A constraint for logical equivalence. An algorithm to reduce each tree to
  some kind of normal form. Not doing this on purpose. Leaving for further work.
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
\item Inference options to explore. Logspace vs normal space. Symbolic vs
  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible. Restarting after each run is
  expensive. Periodic restarts could be an option.
\end{itemize}

@@

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{arguments}[i] = |\constants{}| +
    |\variables{}| - 1.
  \]
\end{constraint}
In other words, we fix the unused arguments to be the last variable in
$\variables{}$ if there are any, and a constant otherwise. Because of the way
constants and variables are ordered, this is the only constant value in $\{ 0,
\dots, |\constants{}| + |\variables{}| - 1 \}$ that satisfies all other
constraints on the head.

% TODO: will need to be updated
\begin{example}
  Let $\maxArity{} = 4$, $\mathsf{P} \in \predicates{}$, $\constants{} = \{ a
  \}$, $\variables{} = \{X, Y, Z \}$. Then $\mathsf{P}(Z, Y, Z)$ would be
  represented as:
  \begin{align*}
    \variable{predicate} &= \mathsf{P},\\
    \variable{arguments} &= [Z, Y, Z, 0],\\
    \variable{arity} &= 3,\\
    \variable{occurrences} &= [\emptyset, \emptyset, \{ 1 \}, \{ 0, 2 \}],\\
    \mathbf{M} &= \begin{bmatrix}
      0 & 0\\
      1 & 1\\
      2 & 0
    \end{bmatrix}.
  \end{align*}
\end{example}

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumClauses{} - 1$, if
  $\variable{headsOfClauses}[i].\variable{predicate} = \Box$, then
  \[
    \variable{bodiesOfClauses}[i].\variable{numNodes} = 1,
  \]
  and
  \[
    \variable{bodiesOfClauses}[i].\variable{treeValues}[0].\variable{name} =
    \top.
  \]
\end{constraint}

\section{Eliminating Variable Symmetries}
% TODO: the matrix should include two more columns for bodies

\begin{definition} We define an array \variable{occurrences} of $|\constants{}|
  + |\variables{}|$ sets such that, for all $t \in \constants{} \cup
  \variables{}$, $\variable{occurrences}[t]$ is a subset of $\{ 0, \dots,
  \maxArity{} - 1\}$ that stores the indices of \variable{arguments} at which
  $t$ is located. It is defined by the \variable{setsIntsChanneling} constraint
  equivalent to the following condition: for $i = 0, \dots, \maxArity{} - 1$ and
  $t \in \constants{} \cup \variables{}$,
  \[
    \variable{arguments}[i] = t \quad \iff \quad i \in \variable{occurrences}[t].
  \]
\end{definition}

\begin{definition}
  We define a $|\variables{}| \times 2$ matrix $\mathbf{M}$ with domain $[0,
  \maxArity{}]$ such that for $v = 0, \dots, |\variables{}| - 1$,
  \begin{align*}
    M_{v,0} &= |\variable{occurrences}[v]|.\\
    M_{v,1} &= \begin{cases}
      \min \variable{occurrences}[v] & \text{if } \variable{occurrences}[v] \ne \emptyset\\
      0\footnotemark & \text{otherwise.}
    \end{cases}
  \end{align*}
\end{definition}
\footnotetext{This could be any fixed value in the domain.}

\section{Interactions Between Clauses}

@@

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$). If
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate},
  \]
  then $\variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i]$.
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body.
% TODO: turn this into lexChainLessEq. But before: update the counting section.

@@

% TODO: does it work? explain why it works
Number of heads for a specific predicate with arity $a \in \arities{}$:
\[
  H_a = \sum_{v = 0}^a |\constants{}|^{a-v} \sum_{k=1}^{\min\{|\variables{}|,
    a\}} p_k(a),
\]
where $p_k(a)$ denotes the number of ways to partition integer $a$ into exactly
$k$ parts.

@@

\begin{definition}[Adjacency matrix representation]
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$
  defined by
  \begin{align*}
    A_{i,j} = 0 \iff &\nexists k: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{name} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{definition}
